---
title: "Homework 03: Use `dplyr`/`ggplot2` to manipulate and explore the `gapminder` data"
author: "**Shuxian Fan**"
date: "Sep 28th, 2018"
output: github_document
always_allow_html: yes
---
## Initial Package Setup
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(gghighlight))
suppressPackageStartupMessages(library(gridExtra))

```

## Exploration Tasks

### (1) Get the maximum and minimum of GDP per capita for all continents
The table shows that Asia has the widest range of GDP per capita values, namely it has more extreme values of GDP per capita. I have highlighted it with a different color. 

I also attached a useful page which has the namded color and hex equivalents for further reference. In order to highlight a certain row, I use the `kableExtra` package, where the function `row_spec` can format certain row you want. 

```{r}
# Using the kable_styling from the kableExtra package to do highlighting.
gap_df1 = gapminder %>%
  group_by(continent) %>%
  summarize(max.gdp = max(gdpPercap), min.gdp = min(gdpPercap))

  knitr::kable(gap_df1,"html")%>%
  kable_styling(bootstrap_options = "striped", full_width = F)%>%    
  row_spec(0, bold = T, color = "black", background = "#E6E6FA" )%>%
  row_spec(3, bold = T, color = "white", background = "darkgrey" )
```
```{r}
gap_df1%>%
  ggplot(aes(continent, max.gdp, fill = continent))+
  geom_bar(stat = "identity")+
  labs(title="Maximum of GDP per capita for continents")


gap_df1%>%
  ggplot(aes(continent, min.gdp, fill = continent))+
  geom_bar(stat = "identity")+
  labs(title="Minimum of GDP per capita for continents")


```

To get a intuitive sense of the maximum and minimum gdpPercap for different continents, I utilize the violin plot along with the jitter plot, where I also add a horizontal line to indicate the maximum values. 
```{r}
summ = 
  gapminder%>%
  group_by(continent) %>%
  summarize(max.gdp = max(gdpPercap), min.gdp = min(gdpPercap))

gapminder%>%
  ggplot(aes(continent, gdpPercap, color = continent))+
  geom_jitter(aes(alpha = 0.4))+
  geom_violin()+
  geom_hline(yintercept = summ$max.gdp, alpha = 0.2)

```

Another way to see the maximum `gdpPercap` is to present how the maximum values change through years. From this plot you also can see the trend.
```{r}
  year_summ = gapminder%>%
  group_by(continent,year) %>%
  summarize(max.gdp = max(gdpPercap), min.gdp = min(gdpPercap))

year_summ%>%
  ggplot(aes(year, max.gdp, shape = continent, color = continent))+
  geom_point()+geom_line()+ 
  labs(title="The Trend of the Maximum `gdpPercap` through years")


year_summ%>%
  ggplot(aes(year, min.gdp, shape = continent, color = continent))+
  geom_point()+geom_line()+ 
  labs(title="The Trend of the Minimum `gdpPercap` through years")
```


### (2) Look at the spread of GDP per capita within the continents

```{r}
summary_stat = 
  gapminder %>%
  group_by(continent) %>%
  summarize(SD = sd(gdpPercap),
            IQR = IQR(gdpPercap),
            Min = min(gdpPercap), 
            Qu_1st = quantile(gdpPercap,0.25), 
            Median = median(gdpPercap), 
            Mean = mean(gdpPercap),
            Qu_3rd = quantile(gdpPercap, 0.75),
            Max = max(gdpPercap)) 
  
knitr::kable(summary_stat,"html")%>%
  kable_styling(bootstrap_options = "striped", full_width = F)%>%    
  row_spec(0, bold = T, color = "black", background = "#E6E6FA" )%>%
  column_spec(1, bold = T, color = "black", background = "#E6E6FA" )
```

```{r}
gapminder%>%
  ggplot(aes(continent, gdpPercap, fill = continent))+
  geom_boxplot(outlier.colour = "red")+
  labs(title="The Boxplots for GDP per capitla of Different Continents")
```

```{r}
theme_set(theme_gray())
gapminder%>%
  ggplot(aes(gdpPercap, fill = continent))+
  facet_wrap(~continent,scale = "free_y")+
  geom_histogram()
```


### (3) Compute a trimmed/weighted mean of life expectancy for different years. 

Based on the definition of Trimmed mean, for most statistical applications, $5-25\%$ of the ends are discarded. The $25\%$ trimmed mean is known as the interquartile mean. Therefore, here I choose to present the interquartile mean of the life expectancy for different years.

```{r}
trimed_df = gapminder %>%
  group_by(year) %>%    # group by year 
  summarize(mean_lifeExp = mean(lifeExp), trimed_mean_lifeExp = mean(lifeExp, trim = 0.25))%>%  
  # summarize the plain mean and trimed mean of lifeExp
  arrange(year)  # arrange by year 

weighted_df = gapminder %>%
  group_by(year) %>%    # group by year 
  summarise(mean_lifeExp = mean(lifeExp), weighted_mean_lifeExp = weighted.mean(lifeExp,pop))%>%
  arrange(year)

 knitr::kable(trimed_df,"html")%>%
  kable_styling(bootstrap_options = "striped", full_width = F)%>%    
  row_spec(0, bold = T, color = "black", background = "#E6E6FA")
 
 knitr::kable(weighted_df,"html")%>%
  kable_styling(bootstrap_options = "striped", full_width = F)%>%    
  row_spec(0, bold = T, color = "black", background = "#E6E6FA")
```

To see how the trimed and weighted mean evolve through years, I used a point plot with line added to illustrate the table above.
```{r}
ggplot(trimed_df, aes(year,trimed_mean_lifeExp))+
  geom_point(color = "red")+
  geom_line(color = "red")+
  geom_point(aes(year,mean_lifeExp))+
  geom_line(aes(year,mean_lifeExp))

ggplot(weighted_df, aes(year,weighted_mean_lifeExp))+
  geom_point(color = "red")+
  geom_line(color = "red")+
  geom_point(aes(year,mean_lifeExp))+
  geom_line(aes(year,mean_lifeExp))
```

### (4) How is life expectancy changing over time on different continents?
The table is too long to present so I made a scroll box using the great feature in the `kableExtra` package.
```{r}
lifeExp_df = gapminder %>%
  group_by(continent, year) %>%
  summarise(mean_lifeExp = mean(lifeExp))
  
knitr::kable(lifeExp_df,"html")%>%
  kable_styling(bootstrap_options = "striped", full_width = F)%>%    
  row_spec(0, bold = T, color = "black", background = "#E6E6FA" )%>%
  scroll_box(width = "400px", height = "400px")
```
To depict how the mean life expectancy change over time, I plot the point along with a linear regression function to capture the trend. 

```{r}

lifeExp_df %>%
  ggplot(aes(year, mean_lifeExp, color = continent))+
  geom_point()+geom_smooth(method = "lm")

# plot the distribution of each continent.
ggplot(gapminder, aes(year,lifeExp))+
  facet_wrap(~continent)+geom_point(alpha = 0.2)+geom_smooth()
  
```

### (5) Report the absolute and/or relative abundance of countries with low life expectancy over time by continent

First, to decide the bench mark of low life expectancey, I choose to take a look at the spread of life expectancy for different continent. From the table below, I decide to choose the mean of first quantiles as the bench mark. After rounding it to the nearest integer, the number is 59.
```{r}
summary_stat_lifeExp = 
  gapminder %>%
  group_by(continent) %>%
  summarize(SD = sd(lifeExp),
            IQR = IQR(lifeExp),
            Min = min(lifeExp), 
            Qu_1st = quantile(lifeExp,0.25), 
            Median = median(lifeExp), 
            Mean = mean(lifeExp),
            Qu_3rd = quantile(lifeExp, 0.75),
            Max = max(lifeExp)) 
  
knitr::kable(summary_stat_lifeExp,"html")%>%
  kable_styling(bootstrap_options = "striped", full_width = F)%>%    
  row_spec(0, bold = T, color = "black", background = "#E6E6FA" )%>%
  column_spec(1, bold = T, color = "black", background = "#E6E6FA" )

benchmark = round(mean(summary_stat_lifeExp$Qu_1st))
```

```{r}
low_lifeExp = 
  gapminder %>%
  group_by(continent, year) %>%
  summarize(low_life_exp_percent =round(sum(lifeExp < 57)/length(lifeExp),2)) %>%
  arrange(year)


  knitr::kable(low_lifeExp,"html")%>%
  kable_styling(bootstrap_options = "striped", full_width = F)%>%    
  row_spec(0, bold = T, color = "black", background = "#E6E6FA" )%>%
  scroll_box(width = "400px", height = "400px") 
  
```
```{r}
ggplot(low_lifeExp, aes(year, low_life_exp_percent))+
  geom_point()+
  geom_line(aes(color = continent))+
   labs(title="Percentage of countries with low life Expectancy over time for all continents")
```

From the plot we can see that, None of the countries in the continent Oceania have low expectancy. Through years, Europe and America no longer have countries with low expectancy. Asia is very close to raise the life expectancy to our benchmark while Africa still have a greater percent of the coutries that have low life expectancy. 

## Reference and Sources

**1. Named Colors and Hex Equivalents**
https://css-tricks.com/snippets/css/named-colors-and-hex-equivalents/

**2. Emoji Cheat Sheet**
https://www.webpagefx.com/tools/emoji-cheat-sheet/

**3. Truncated Mean**
https://en.wikipedia.org/wiki/Truncated_mean

**4. Create Awesome HTML Table with knitr::kable and kableExtra**
https://haozhu233.github.io/kableExtra/awesome_table_in_html.html